---
title: "MB5370 Module 04. Assignment 2 - QFISH"
author: "Alexa Hoffman"
output: html_document
date: "2026-04-26"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This assignment will focus on start to finish data cleaning to visualisation using QFISH data from the Queensland Government. I will be working with the Commercial Harvest dataset that focuses on the Tropical Rock Lobster fishery.

### Data Loading
```{r message=FALSE, warning=FALSE}
library(tidyverse)

# Load data
harvest_data <- read.csv("/Users/alexa/Desktop/Data_Analysis/MB5370_Techniques/MB5370_Module_4_Hoffman/data/QFISH_Lobster_Harvest.csv")

# View raw data structure
head(harvest_data)
tail(harvest_data)

#reload in data with treating first row as data
raw_data <- read.csv("/Users/alexa/Desktop/Data_Analysis/MB5370_Techniques/MB5370_Module_4_Hoffman/data/QFISH_Lobster_Harvest.csv", 
                header = FALSE, stringsAsFactors = FALSE)
```

### Tidying the dataset
Need to select out just the tropical lobster data, removing 'cray tails' and 'unspecified'. Additionally, the first 5 lines of data are to be removed as they are invalid or metadata.
```{r}
lobster <- raw_data[5:nrow(raw_data), c(1, 6, 7, 8)]
colnames(lobster) <- c("Year", "Licences", "Days", "Tonnes")

# add species column
lobster$Species <- "Tropical Rock Lobster"

# clean out NAs and invalid data
tidy_lobster <- lobster |>
    filter(
      !Year %in% c("N/A", "Grand Total"),
      !Licences %in% c("N/A", ""),
      !Days %in% c("N/A", ""),
      !Tonnes %in% c("N/A", "")
    )
# check cleaned data
head(tidy_lobster)
tail(tidy_lobster)

# view summary, checking variables
summary(tidy_lobster)
```

These data columns are currently characters, they need to be transformed to numerical objects. 
```{r}
#switching character columns to numeric
tidy_lobster <- tidy_lobster %>%
  mutate(
    Year = as.numeric(Year),
    Licences = as.numeric(Licences),
    Days = as.numeric(Days),
    Tonnes = as.numeric(Tonnes)
  )

summary(tidy_lobster) #fixed!
```

### Data Visualisation
For plotting these data I am looking at the total catch from the commercial harvest fishery for the tropical lobster fishery over time. This will allow to see fluctuations in total catch. 
```{r}
# Plot
ggplot(data = tidy_lobster, mapping = aes(x = Year, y = Tonnes)) +
  geom_line(color = "darkblue", size = 0.5) +
  labs(
    title = "Total catch (tonnes) of Tropical Rock Lobster",
    subtitle = "From commerical harvest fisheries in Queensland, 1995 to 2025",
    x = "Year",
    y = "Catch (Tonnes)",
    caption = "Source: QFISH commercial harvest data, accessed: April. 25, 2026."
  ) +
  scale_x_continuous(breaks = seq(min(tidy_lobster$Year, na.rm = TRUE), 
                                  max(tidy_lobster$Year, na.rm = TRUE), 
                                  by = 2)) +
  theme_bw()

ggsave("rocklobster_timeseries_plot.png", width = 10, height = 6, dpi = 300)
```
Extended caption: Annual commercial harvest of Tropical Rock Lobster in Queensland shows high inter-annual variability, rising from 1995 with peak catches reaching 243.89 tonnes in 2008 and a declining trend observed from 2022 to 2024.