---
title: "MB5370 Module 04. Workshop 4 - Spatial data in R"
author: "Alexa Hoffman"
output: html_document
date: "2026-04-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# 5.3 R and how we use it for spatial analysis
R can also be used as a Geographic Information System (GIS), capable of doing most things that ArcGIS or QGIS can do — and for free.
The key advantages of doing spatial analysis in R are:

  - Your workflow is written as a script, making it easy to reproduce, improve and share
  - You can version control your analysis using Git and GitHub
  - Results can feed directly into other R tools like ggplot2, allowing you to build complete end-to-end analyses in a single script
  - Scripts can be shared openly, supporting reproducible and transparent science

As with everything else we have covered, we will be working in the tidyverse to keep our spatial data tidy and ready for analysis.

# 5.4 Installing the spatial R packages
This section will use a dataset `copepods_raw.csv`
```{r message=FALSE, warning=FALSE, results='hide'}
#Packages
library(tidyverse)
library(sf)
library(terra)
library(leaflet)
library(tmap)
library(mgcv)
```
# 5.5 Introduction to the problem
Here we adapt A/Prof Brown’s statement of the problem. 

  *You finally have a chance to meet one of your academic heroes. On meeting her, she mentions that she’s read your first PhD paper on zooplankton biogeography. She said she was particularly impressed with the extent of R analysis in your biogeography paper and goes on to suggest you collaborate on a new database she is ‘working with’.*
  *The database has extensive samples of copepod richness throughout Australia’s oceans and the Southern Ocean too. She has a hypothesis - that like many organisms, copepod species richness (which is the number of unique species) will be higher in warmer waters than cooler waters. But she needs help sorting out the data.*
  *First and foremost, she wants you to use your skills in R to help develop a map that could help you ‘get a look at’ whether this hypothesis is worth pursuing.* 

**Your task now is - using R - to make a map of copepods in relation to temperature.**

# 5.6 Downloading and loading the spatial dataset

  *Your hero professor has sent you the data files in a .zip format. The spreadsheet copepods-raw.csv has measurements of copepod species richness from around Australia. Copepods are a type of zooplankton, perhaps the most abundant complex animals on the planet and an important part of ocean food-webs. *
 *Like many distracted academics, she has also sent you some other data, but has not explained what it is all for yet. You’ll have to figure that out.*
  *Copepod species were counted using samples taken from a Continuous Plankton Recorder. The CPR was towed behind ‘ships of opportunity’ (including commercial and research vessels). ‘Silks’ run continuously through the CPR and the plankton are trapped onto the silks, kind of like a printer that runs all day and night to record plankton in the ocean.*
  *So these data are what we’ll work with today. As a realistic learning experience, be ready to face some errors in the data received from our distracted professor*!
  
  Starting with that copepod richness data. In this part of the course we are going to clean it up and run some basic analyses.
```{r message=FALSE, warning=FALSE}
library(readr)
dat <- read_csv("/Users/alexa/Desktop/Data_Analysis/MB5370_Techniques/MB5370_Module_4_Hoffman/data/data-for-course/copepods_raw.csv")
dat
```
# 5.7 Data exploration
Since we don’t know this copepod data well and we haven’t been told much about it (and it lacks any metadata on what it all means), we need to thoroughly check it and make sure we understand it.

## *5.7.1 Check coordinates*
Step 1: make a map in `ggplot2`, plotting the coordinates for the samples (segments of the CPR silks)
```{r message=FALSE, warning=FALSE}
library(ggplot2)
ggplot(dat) + 
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point()
```
This plot shows the location of every segment, coloured by species richness. Notice that the x and y axes are longitude and latitude — essentially giving us a simple map, since ggplot just plots points on an x-y grid.
But this is not a true map, it doesn't have a projection, therefore the real distances between these points when measured with a ruler are probably wrong.

Looking at the richness in a non-spatial way. Latitude on the x-axis and richness on the y.
```{r message=FALSE, warning=FALSE}
ggplot(dat, aes(x = latitude, y = richness_raw)) + 
  stat_smooth() + 
  geom_point()
```
By visualising your data it will help you understand it and be able to ask questions. In this example there is clearly a shift in this graph around latitude -40. This is when you would to ask your collaborator questions.

# 5.8 Getting going with maps
Repeating the map of richness, now using R spatial packages. 

First, we will turn our point data into a spatially referenced data frame using the sf package (sf stands for ‘simple features’) which is an open standard for geospatial databases. For those that think in GIS, you can think of this format as a shapefile or feature collection.

Turning data into a 'simple features collection`
```{r message=FALSE, warning=FALSE}
?st_as_sf #good practice to look at functions and packages
sdat <- st_as_sf(dat, coords = c("longitude", "latitude"), 
                 crs = 4326)
```

  - `st_as_sf` converts different data types to simple features. 
dat is our original data. 
  - `coords` gives the names of the columns that relate to the spatial coordinates (in order of X coordinate followed by Y coordinate).
  - `crs` stands for coordinate reference system
  
# 5.9 Coordinate reference systems
In short, coordinate reference systems are required for 2D mapping to compensate for the lumpy, spherical (3D) nature of the earth.

*In mapping, a datum defines the reference point and an ellipsoid models the shape of the earth. Together they form a Geographic Coordinate System (GCS), which tells R where our coordinates are located on the earth. Coordinates are expressed as longitude and latitude in decimal degrees.*

*For our copepod data we use a common GCS called WGS84, referenced by its EPSG code 4326. An EPSG code is simply a short unique code for a specific coordinate reference system.*

*In R, best practice is to define a CRS using either an EPSG code or a Well-Known Text (WKT) string. WKT contains all the detailed information about a CRS but EPSG codes are simpler when you don't need that level of detail.*

```{r message=FALSE, warning=FALSE}
crs4326 <- st_crs(4326)
crs4326 # look at the whole CRS
crs4326$Name # pull out just the name of the crs

crs4326$wkt # crs in well-known text format
```

*When we make a 2-dimensional map in WGS84 GCS, we assume that a degree is a linear unit of measure (when in reality it’s angular).*

*To more accurately map our data in 2 dimensions, we need to decide how to ‘project’ 3 dimensions into 2.* 

*There are many ways to do the projection depending on where we are in the world and what we’re most interested in preserving (e.g., angles vs. distances vs. area). Projections are defined by a projected coordinate reference system (PCS), and spatial packages in R use the software PROJ to do this.*

To find the most appropriate projected crs for your data, try the R package `crssuggest`.

# 5.10 Feature collection (points)
Looking back at what we created:
```{r message=FALSE, warning=FALSE}
sdat
```
*The data table in sdat looks much like dat did, but note it now has a geometry column. This is where the coordinates (just one point for each data row) are stored. More complex simple features could have a series of points, lines, polygons or other types of shapes nested in each row of the geometry column.*

*The nice thing about sf is that because the data is basically a dataframe with a geometry, we can use all the operations that work on dataframes on sf simple features collections.*

*These include data wrangling operations like inner_join, plotting operations from ggplot2 and model fitting tools too (like glm).*

*sf also adds geometric operations, like st_join which do joins based on the coordinates.*

In summary, a simple feature is like a shapefile, in that it holds a lot of data in columns and rows but is spatially aware. Essentially, that includes extra columns regarding each row's position (in coordinates) and metadata about the coordinate reference system, the type of geometry (Point) and so on.

# 5.11 Cartography
`sf` has simple plotting features
```{r message=FALSE, warning=FALSE}
plot(sdat["richness_raw"])
```

Here we have only plotted the richness column. If we used `plot(sdat)` it would create a panel for every variable in our dataframe. In sf, we can use square brackets [`"richness_raw"`] to select a single variable.
```{r message=FALSE, warning=FALSE}
plot(sdat)
```

# 5.12 Thematic maps for communication
In this module we will use `tmap`. `tmap` works similarly to ggplot2 in that we build and add on layers. Here we only have one layer from `sdat`. We declare the layer with `tm_shape()` (in this case `sdat`), then the plot type with the following command.

```{r message=FALSE, warning=FALSE}
#using tmap
tm_shape(sdat) + 
  tm_dots(col = "richness_raw")
```
Customizing:
Adding in a color scale to display richness as a continuous varible.
```{r}
library(viridis)
tm1 <- tm_shape(sdat) +
  tm_dots(fill = "richness_raw", fill.scale = tm_scale(values = "viridis"))

tm1
```

```{r}
tmap_save(tm1, filename = "Richness-map.png", 
          width = 600, height = 600,
          device = png)
```

# 5.13 Mapping spatial polygons as layers

`sf` package can handle many types of spatial data, including shapes like polygons. To practice with polygons we will load in a map of Australia and a map of Australia’s continental shelf using `tmap` to add these layers.

## *5.13.1 Loading shapefile*
Unlike our previous csv data with coordinate columns, the copepod polygon data is stored as a shapefile (.shp).

Shapefiles are considered an outdated format — they are inefficient and actually consist of multiple files, meaning data can easily be lost if files are transferred incomplete. A better alternative is the Geopackage (.gpkg) format, which stores everything in a single compressed file.

We are using shapefiles here because they are still the most common format you will encounter in practice, but it is good practice to save your own spatial data as .gpkg going forward.

Shapefiles can be read into R using `st_read()`, which works like `read_csv()` but for spatial files.

```{r}
aus <- st_read("/Users/alexa/Desktop/Data_Analysis/MB5370_Techniques/MB5370_Module_4_Hoffman/data/data-for-course/spatial-data/Aussie/Aussie.shp")

aus #already spatially aware
```

```{r}
shelf <- st_read("/Users/alexa/Desktop/Data_Analysis/MB5370_Techniques/MB5370_Module_4_Hoffman/data/data-for-course/spatial-data/aus_shelf/aus_shelf.shp")

shelf #already spatially aware
```

## *5.13.2 Mapping your polygons*
`tmap` makes it very straightforward to make a map of polygons:
```{r}
tm_shape(shelf) + 
  tm_polygons()
```

Remember we can make a thematic map by layering it up just as we do for plots in ggplot2. 

To add more layers:
```{r}
tm_shape(shelf, bbox = sdat) + 
  tm_polygons() +
  tm_shape(aus) + 
  tm_polygons() + 
  tm_shape(sdat) + 
  tm_dots()
```

  - Here we have indicated the shape of our map (`shelf`) and we have added a command `bbox = sdat` to expand the extent of the map so it depicts all of our copepod data points. 
  - We then add the shape of Australia (aus) on top of the shelf, and 
  - finally our copepod data (sdat) in the form of points using tm_dots().

## *5.14 Exploring t_map*

Running the exercises from [here](https://r-tmap.github.io/tmap/articles/tmap-getstarted.html)

```{r}
library(tmap)
## Breaking News: tmap 3.x is retiring. Please test v4, e.g. with
## remotes::install_github('r-tmap/tmap')
data("World")

tm_shape(World) +
    tm_polygons("HPI")

?'tmap-element'
```
The `World` object is an `sf` spatial dataframe where each row has an associated polygon geometry. To plot it in `tmap` you first specify the data with `tm_shape()`, then add layers using `+` — in this case `tm_polygons()`. All `tmap` layer functions use the `tm_` prefix, making them easy to find in the documentation.

**Interactive Maps**
Each map can be plotted as a static image or viewed interactively using "plot" and "view" modes, respectively. The mode can be set with the function `tmap_mode`, and toggling between the modes can be done with the ‘switch’ `ttm()` (which stands for toggle thematic map).
```{r}
tmap_mode("view")

tm_shape(World) +
    tm_polygons("HPI")
```
This allows for zooming in and moving around the map.

**Multiple shapes and layers**
A shape is a spatial object (with a class from `sf`, `sp`, `stars`, or `raster`). Multiple shapes and also multiple layers per shape can be plotted:
```{r}
data(World, metro, rivers, land)

tmap_mode("plot")
## tmap mode set to plotting
tm_shape(land) +
    tm_raster("elevation", col.scale = tm_scale(values = terrain.colors(10))) +
tm_shape(World) +
    tm_borders("white", lwd = .5) +
    tm_text("iso_a3", size = "AREA") +
tm_shape(metro) +
    tm_symbols(col = "red", size = "pop2020", size.scale = tm_scale_continuous(values.scale = .5)) +
tm_legend(show = FALSE)
```
Note: had to update the code using the suggestions R gave because some of the arguments were for the wrong version of tmap.

**Facets**
Facets can be created in three ways:

  1. By assigning multiple variable names to one aesthetic (in this example the first argument of `tm_polygons`):
```{r}
tmap_mode("view")
tm_shape(World) +
    tm_polygons(c("HPI", "economy")) +
    tm_facets(sync = TRUE, ncol = 2)
```

  2. By splitting the spatial data with the by argument of `tm_facets`:
```{r}
tmap_mode("plot")
## tmap mode set to plotting

data(NLD_muni)

names(NLD_muni)

tm_shape(NLD_muni) +
  tm_polygons("employment_rate",
              fill.scale = tm_scale(values = "brewer.rd_yl_bu")) +
  tm_facets(by = "province")
#had to alter the varible being looked at bc 'pop_men' didn't exist in this version of the data
```

  3. By using the tmap_arrange function:
```{r}
tmap_mode("plot")
## tmap mode set to plotting

data(NLD_muni)
tm1 <- tm_shape(NLD_muni) + tm_polygons("population", convert2density = TRUE)
tm2 <- tm_shape(NLD_muni) + tm_bubbles(size = "population")

tmap_arrange(tm1, tm2)
```
**Basemaps and overlay tile maps**
Tiled basemaps can be added with the layer function `tm_basemap`. Semi-transparent overlay maps (for example annotation labels) can be added with `tm_tiles`.

```{r}
tmap_mode("view")
tm_basemap("Stadia.StamenWatercolor") +
tm_shape(metro) + tm_bubbles(size = "pop2020", col = "red") +
tm_tiles("Stadia.StamenTonerLabels")
```

**Options and styles**
The functions `tm_layout` and `tm_view` are used to specify the map layout and the interactive aspects respectively. These functions can be used in the same way as the layer functions, e.g.
```{r}
tmap_mode("plot")
## tmap mode set to plotting

tm_shape(World) +
    tm_polygons("HPI") +
tm_layout(bg.color = "skyblue", inner.margins = c(0, .02, .02, .02))
```

These options, as well as a couple of others, can also be set within with `tmap_options`, which works in the same way as the base R function options. The main advantage is that these options are set globally, so they do not have to be specified in each map, for the duration of the session.

```{r}
tmap_options(bg.color = "black", legend.text.color = "black")

tm_shape(World) +
  tm_polygons("HPI", fill.legend = tm_legend(title = "Happy Planet Index"))
```
A style is a certain configuration of the tmap options.

```{r}
tmap_style("classic")
## tmap style set to "classic"
## other available styles are: "white", "gray", "natural", "cobalt", "col_blind", "albatross", "beaver", "bw", "watercolor"

tm_shape(World) +
    tm_polygons("HPI", fill.legend = tm_legend(title = "Happy Planet Index"))
```

```{r}
# see what options have been changed
tmap_options_diff()
## current tmap options (style "classic") that are different from default tmap options (style "white"):
## $frame.double_line
## [1] TRUE

## $color.sepia_intensity
## [1] 0.7

## $text.fontfamily
## [1] "serif"

## $inset.type
##[1] "rose"

# reset the options to the default values
tmap_options_reset()
```
New styles can be created; see `?tmap_options`.

**Exporting maps**
```{r}
tm <- tm_shape(World) +
    tm_polygons("HPI", fill.legend = tm_legend(title = "Happy Planet Index"))

## save an image ("plot" mode)
tmap_save(tm, filename = "world_map.png")

## save as stand-alone HTML file ("view" mode)
tmap_save(tm, filename = "world_map.html")
```

**Shiny integration**
It is possible to use tmap in shiny:
```{r eval=FALSE}
# in UI part:
tmapOutput("my_tmap")

# in server part
output$my_tmap = renderTmap({
    tm_shape(World) + tm_polygons("HPI", tm_legend(title = "Happy Planet Index"))
})
```

**Quick thematic map**
Maps can also be made with one function call. This function is `qtm`:
```{r}
qtm(World, fill = "HPI", 
    fill.scale = tm_scale(values = "brewer.rd_yl_gn")) #updated with v4 syntax
```

**Tips n Tricks**
```{r}
tmap_tip()
```