---
title: "MB5370 Module 04. Workshop 2 - Using ggplot for communication"
author: "Alexa Hoffman"
output: html_document
date: "2026-03-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# load tidyverse packages
```{r}
library("tidyverse")
```

# Load the data
```{r message=FALSE, warning=FALSE, results='hide'}
data(mpg)
```


# Quick data checks
```{r message=FALSE, warning=FALSE, results='hide'}
head(mpg)
glimpse(mpg)
summary(mpg)
```


# 3.1 Labels
In ggplot2, you add labels with the labs() function.
```{r message=FALSE, warning=FALSE}
#Title
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se.e = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size")
```
For more text, can add in other functions:
  - 'subtitle'
  - 'caption'
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )
```
labs() can also be used to replace axis labels and legend titles
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    colour = "Car type"
  )
```

#3.2 Annotations
When you want to add text to the plot itself. You can use 'geom_text()' to add text labels to the plots. This works the same way as 'geom_point', just in text vs shape form.
```{r}
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class)
```
This filters the data and adds a label that calls in the values from the dataframe. In this case it helps to pull out the most fuel efficient cars and label it within the plot.

When labels overlap (like above) it can be adjusted with the 'nudge()' function
or other packages for text wrapping.

They can be good tools for highlighting data such as min and max.

#3.3 Scales
There are automated scales that exist in defult in ggplot2. 
Ex:
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))
```
To provide limits you’ll need to offer a character vector showing the start and end of your limit (e.g. c(0,12)). Note also that: ?scale_x_continuous() has plenty of other arguments, so don’t forget to explicitly state the argument you’re providing these limits for (e.g. limits = c(0,12)). 

```{r}
?scale_x_continuous() 

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous(limits = c(3,6)) +
  scale_y_continuous(limits = c(10,30)) +
  scale_colour_discrete()
```





