---
title: "MB5370 Module 04. Workshop 3 - Data Wrangling"
author: "Alexa Hoffman"
output: html_document
date: "2026-03-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



#4.2 Workshop Overview - Objectives
This workshop will focus on data tidying using `tidyr`:
- Learn the principles of tidy data using tidyr
- Learn how to import datasets into R
- Learn how to join data with dplyr. 

Tibbles: are essentially dataframes, helps to handle issues of aging programming language. Aka 'future proof' dataframes.

#4.3 Tidying data using Tidyr
Tidyr is a part of the larger tidyverse package. By creating tidy data it allows for more tie analyzing data and less manipulating it
load tidyverse package:
```{r message=FALSE, warning=FALSE}
library("tidyverse")
```

#4.4 Tidy data
There are many ways to display a given data set, but not every way is easy to use for analysis. For example, the format of a field datasheet might be convenient for data collection and entry, but it might not be a useful way to display the data when you go to analyse it. The process of rearranging your data to a format more appropriate for analysis is the process of “tidying.”

EX:
```{r}
table1
```
```{r}
table2
```
```{r}
table3
```
All three of these tables display the same data, but only `table1` expresses 'tidy' data. These data follow the... 
**Three interrelated rules of tidy data:**
*1. Each variable must have its own column.*
*2. Each observation must have its own row.*
*3. Each value must have its own cell.*

These three rules are interrelated because it’s impossible to only satisfy two of the three. That interrelationship leads to an even simpler set of practical instructions:
  *1. Put each dataset in a tibble (special dataframe)*
  *2. Put each variable in a column.*

Now, why do we care about having tidy data? We said before, tidy data is useful data and here’s why: 
  *1. Having a consistent data structure makes it easier to learn the tools that work with it, and* 
  *2. Having your variables in columns allows R to use its ability to work with vectors of values. This makes transforming tidy data a smoother process.*

All packages in the tidyverse are designed to work with tidy data. Including ggplot2.

**Pipes**
A pipe is really only designed to help you better understand what the code is doing. It takes the data (left of the pipe) and applies the function (right of pipe). I’ll use both `%>%`, and `|>`  which achieve the exact same thing (|> is brand new in base R, %>% only works in tidyr and magrittr packages)
```{r}
#Compute rate per 10,000
table1 %>%
  mutate(rate = cases / population * 10000)

#Compute rate per year
table1 %>%
  count(year, wt = cases)

#Visualise changes over time
library(ggplot2)
ggplot(table1, aes(year, cases)) +
  geom_line(aes(group = country), colour = "grey50") +
  geom_point(aes(colour = country, shape = country))
```
##4.4.1 Exercises
1. For each of the sample tables, describe what each observation and each column represents.
`table1` 
- Each observation is for a country, where each column respresents a separate variable including: a single year, number of cases for that year in that country, and population of that country on the given year. 
`table2`
- Each observation is a country, but not every column is a variable. Each year the given country has two observations, one for the number of cases and the second for the population. The columns then become more confusing as they are 'type' and 'count' with type representing the variables of cases and population, and the 'count' as the numerical value of those variables. 
`table3`
- Each observation is a country and year, but then the columns do not represents the variables. Instead for each year in the country the cases per population is expresses as a rate, expressed as a fraction. The only way you know its cases per population is if you look back at table1. Rate could be a variable in its own column, but would need to be calculated from two separate columns with cases and population as separate variables for it to be expressed as true tidy data.Additionally rate needs to be expressed as a frequency, so you would need to multiply the fraction against your unit of measure. Ex: cases/pop * 10,000 (cases of TB per 10,000 people in that given population)

2. Sketch out the processes you would use to calculate the `rate` for `table2` and `table3`. You will need to perform four operations:

`table2`
Because cases and population are stored in the same column (count) separated by a type label, you would first need to split the data into two separate tables, one containing only the rows where type = "cases" and one where type = "population". Then match them into one table by country and year so each row has both values side by side (like table 1). Then divide cases by population and multiply by 10,000. Then store (mutate) the result as a new column called `rate` alongside country and year.

`table3`
The rate column already contains both pieces of information but stored as a single character string (e.g. "745/19987071") rather than two separate numbers. You would need to split that string at the "/" to separate the numerator (cases) and denominator (population) into two new numeric columns. Then divide cases by population and multiply by 10,000. Then store the result as a new numeric rate column, replacing the original.

#4.5 Pivoting data to make it tidy
The first step in tidying the data is to understand what each variable and observation actually means. Sometimes this is obvious and sometimes you’ll need to consult with the person(s) who collected the data. By understanding tidy data structure, you streamline input of data in the future. 

Once you understand the data you’re looking at, the second step is to resolve one of the two common problems with untidy data. These are:
1. One variable is spread across multiple columns
2. One observation is scattered across multiple rows

To fix these we will **pivot** our data (i.e. move it around) into tidy form using two functions in `tidyr`: `pivot_longer()` to lengthen data and `pivot_wider()` to widen data. 

##4.5.1 Lengthening datasets
Pivoting data to be longer is the most common tidying issue that is faced. `pivot_longer()` makes datasets “longer” by increasing the number of rows and decreasing the number of columns, solving those common problems of data values in the variable name (e.g wk1, wk2, wk3, etc.).

Let’s see this in action by using `pivot_longer()` in an example given by R4DS:
The billboard dataset records the billboard rank of songs in the year 2000.

```{r}
billboard
```
In this dataset, each observation is a song. The first three columns (artist, track and date.entered) are variables that describe the song. Then we have 76 columns (wk1-wk76) that describe the rank of the song in each week. Here, the column names are one variable (the week) and the cell values are another (the rank). To tidy the billboard dataset we will use pivot_longer().

This is the case where actual data values (wk1, wk2 etc.)  are in the column name, with each observation (row of data) being a song. We need to have the data in a format where each row is an observation (so-called long format).
```{r}
billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank"
  )
```
*Three key arguments to the `pivot_longer()` function:*
- `cols` which specifies the columns you want to pivot (the ones that aren’t variables). Note: you could either use `!c(artist, track, date.entered)` OR `starts_with('wk')` because the cols argument uses the same syntax as `select()`.
- `names_to` which names the variable stored in the column names. We chose to name that variable `week`.
- `values_to` which names the variable stored in the cell values that we named `rank`

Note that in the code `"week"` and `"rank"` are quoted because they are new variables that we are creating, they don’t exist yet in the data when we run the `pivot_longer()` call.

There are a number of NAs in the above output. It looks like “Baby Don’t Cry” by 2 Pac was only in the top 100 for 7 out of 76 weeks. Therefore, when we lengthened the data, the weeks where it wasn't on the charts became ‘NA.’ These NA’s were forced to exist because of the structure of the dataset not because they are actually unknown. Therefore, we can simply ask `pivot_longer` to remove them by adding the argument `values_drop_na = TRUE` as shown below:
```{r}
billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  )
```

##4.5.2 Pivoting longer
We’ve just seen how pivoting can help reshape our data but let's look further into what pivoting actually does to our data. We will start with a simple dataset and once again follow an example from R4DS. Note the use of the term “tribble” here (not the same as tibble) but also a type of dataframe that allows us to construct small tibbles by hand. Don’t get too caught up in this, simply follow along with the example.
```{r}
df <- tribble(
  ~id, ~bp1, ~bp2,
  "A", 100, 120,
  "B", 140, 115,
  "C", 120, 125
)
```
We want our new (tidy) dataset to have 3 variables:
1. `id` which already exists
2. `measurement` the column names
3. `value` the cell values

To make this happen the pivot longer function needs to be used on df:
```{r}
df |>
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
```
For the visualization of how this data transformed, refer to the module workbook and [Section 5.3.3 in the textbook](https://r4ds.hadley.nz/data-tidy#many-variables-in-column-names).

##4.5.3 Widening datasets
In less common cases, we may need to widen a dataset rather than lengthen it. Widening is essentially the opposite of lengthening and we do so by using the function `pivot_wider()`. `pivot_wider()` allows us to handle an observation if it is scattered across multiple rows. Example would be `table2` -> `table1`
In table2 we see an observation is a country in a year with the observation spread across two rows (cases or population and their values). The table needs to be made wider to move the value of cases into one column, and those of population into another column to comply with the 3 rules of tidy data.

EX:
```{r}
cms_patient_experience
```
The core unit being studied is an organization. But in this format, each organization is spread across six rows with one row for each measurement taken in the survey organization. We can see the complete set of values for `measure_cd` and `measure_title` by using `distinct()`:
```{r}
cms_patient_experience |>
  distinct(measure_cd, measure_title)
```
Neither of these columns will make particularly great variable names: `measure_cd` doesn’t hint at the meaning of the variable and `measure_title` is a long sentence containing spaces. `measure_cd` will be used as the source for the new column names for now, but in a real analysis should create own variable names that are both short and meaningful.

`pivot_wider()` has the opposite interface to `pivot_longer()`: instead of choosing new column names, we need to provide the existing columns that define the values (`values_from`) and the column name (`names_from`):
```{r}
cms_patient_experience |>
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )
```
The above output doesn’t look quite right; we still seem to have multiple rows for each organization. That’s because, we also need to tell `pivot_wider()` which column or columns have values that uniquely identify each row; in this case those are the variables starting with "`org`":
```{r}
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
```
This gives the output that we want.

##4.5.4 Pivoting wider
To understand what pivot_wider() does to our data, let’s once again use a simple example. This time we have two patients with id s A and B, and we have three blood pressure (bp) measurements from patient A and two from patient B:
```{r}
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)
```

The names will be taken from the `measurement` column using the `names_from()`, abd the values from the `value` column using the `values_from()`:
```{r}
df |>
  pivot_wider(
    names_from = measurement,
    values_from = value
  )
```

**Below breaks down how the function actually achieves this, step by step:**

To start the pivoting process, pivot_wider() needs to first figure out what will go in the rows and columns. The new column names will be the unique values of measurement.
```{r}
df |> 
  distinct(measurement) |> 
  pull()
```

When `pivot_wider()` is deciding what the rows should be, it looks at whichever columns aren't being used to make the new column names or fill in the values. In this case that's just the id column, so each unique patient becomes one row. These leftover columns are called `id_cols`. In a more complex dataset there could be several of these — for example country and year — and each unique combination of them would become its own row.
```{r}
df |> 
  select(-measurement, -value) |> 
  distinct()
```

pivot_wider() then combines these results to generate an empty dataframe
```{r}
df |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA)
```

## 4.5.5 Exercises
### 1. Why are pivot_longer() and pivot_wider() not perfectly symmetrical? Carefully consider the following example. 

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)

stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
 pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return")
```
(Hint: look at the variable types and think about column names) pivot_longer() has a names_ptypes argument, e.g.  names_ptypes = list(year = double()). What does it do?

When the data were formatted into the wider format, the numerical values of year were converted to a column name, switching the year to a text/character. Then when it was pulled back into the long format with `pivot_longer` it was treated as a character, this created the asymmetry.

`names_ptypes` is an arugement in `pivot_longer` that allows you to confirm with R what data type your columns are when it is pulled back in. To actually tell R to transform it to a certain type you need to use the `names_transform` function.

`names_ptypes = list(year = double())` tells R to confirm the year column as a number and not text.


