---
title: "MB5370 Module 04. Workshop 1 - Introduction"
author: "Alexa Hoffman"
output: html_document
date: "2026-03-05"
editor_options: 
  markdown: 
    wrap: sentence
---
This is the second part of workshop 1, from 2.22 forward. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Install and load tidyverse packages

```{r message=FALSE, warning=FALSE, results='hide'}
library("tidyverse")
```

# Load the data (mpg is built into ggplot2)

```{r message=FALSE, warning=FALSE, results='hide'}
data(mpg)
```

# Quick data checks

```{r message=FALSE, warning=FALSE, results='hide'}
head(mpg)
glimpse(mpg)
summary(mpg)
```

# 2.22 Transformations and Stats

### Load data

```{r message=FALSE, warning=FALSE, results='hide'}
data("diamonds")
glimpse(diamonds)
```

### Bar plot

```{r message=FALSE, warning=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

On the x-axis, the chart displays cut, a variable from the diamonds dataset.
On the y-axis, it displays count, but count is not a variable in diamonds!
Many graphs, like scatterplots, plot the raw values of your dataset.
Other graphs, like bar charts, calculate new values to plot: - bar charts, histograms, and frequency polygons “bin” your data and then plot bin counts (the number of points that fall in each bin).
- smoothers fit a model to your data and then plot predictions from the model.
- boxplots compute a robust summary of the distribution and then display a specially formatted box.
The algorithm used to calculate new values for a graph is called a stat, short for statistical transformation.

You can generally use geoms and stats interchangeably.
For example, can recreate the previous plot using 'stat_count'.

```{r message=FALSE, warning=FALSE}
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```

## 22.22.2 Overriding defaults

You should always learn about the given defaults, as it can pose implications for your data.

Make some new data

```{r message=FALSE, warning=FALSE, results='hide'}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo
```

```{r message=FALSE, warning=FALSE}
ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```

**Question: Why did we specify now a specific y axis?** We specified a specific y axis because we are moving past the default of count on the y-axis and instead using absolute frequency.

### Proportions rather than freq

What if we want to know the proportion rather than the absolute frequency?

```{r message=FALSE, warning=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop), group = 1))
```

## 2.23.3 Plotting statistical details

You might also want to show a little more about these transformations in your plot, which is good practice to be transparent about uncertainty or any other limitation of your data.
You can do this using stat_summary().

```{r message=FALSE, warning=FALSE}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```

# 2.23 Aesthetic adjustments adjustments

Another way to boost the way you can convey information with plots using ggplot2 is to use aesthetics like colour or fill to change aspects of bar colours.
We already did this once, but there are multiple options available to you, including changing the fill our outline colours.

```{r message=FALSE, warning=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```

**Question: Does anyone notice anything different in the colour scheme of this plot?** (Hint: It's in the viridis colour palette (colour blind friendly), but why is it different from the colour palette we used earlier?)
This scheme uses more of a gradient than the palette we used before.
It is a perceptually uniform sequential scale, which aids in the understanding that the diamonds are graded on a scale with the various labels as categories they are split into.
Whereas, before the color palette used for the cars were split up by type of vehicle, which was appropriate because these car types are not on a scale, but distinct different entities.
Check out the difference:

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) 

ggplot(data = diamonds, mapping = aes(x = color, y = depth)) + 
  geom_point(mapping = aes(color = cut)) 
```

### Filling by a variable

Now try using these aesthetics to colour by another variable like clarity.
Notice how the stacking is done automatically.
This is done behind the scenes with a position argument.

```{r message=FALSE, warning=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```

##Position adjustments The ability to make position adjustments is vital, it allows you to customise your plots in three ways, identity (raw data), fill (changes heights) and dodge (which forces ggplot2 to not put things on top of each other)

### position = "identity"

You will be able to place each object exactly where it falls in the context of the graph.

This is vital for point charts like scatter plots but makes a mess in a bar plot situation by showing too much information (a bar plot generally summarises information).
So in this case we will need to alter the bar aesthetic.

#### To color the bar outlines with no fill color

```{r message=FALSE, warning=FALSE}
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```

### To alter transparency (alpha)

```{r message=FALSE, warning=FALSE}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")
```

### position = "fill"

Works like stacking, but makes each set of stacked bars the same height.

```{r message=FALSE, warning=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```

### position = "dodge"

Places overlapping objects directly beside one another.

```{r message=FALSE, warning=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

# Jittering

position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap.
This is useful for scatterplots but not barplots.

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
  #geom_point(mapping = aes(x = displ, y = hwy))
```

# 2.24 The layered grammar of graphics

The foundation to make almost any plot in ggplot2:

```{r message=FALSE, warning=FALSE, eval=FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <FACET_FUNCTION>
```
