---
title: "MB5370 Module 04. Workshop 1 - Introduction"
author: "Alexa Hoffman"
output: html_document
date: "2026-03-05"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r message=FALSE, warning=FALSE, results='hide'}
library("tidyverse")
```

# Load the data (mpg is built into ggplot2)

```{r message=FALSE, warning=FALSE, results='hide'}
data(mpg)
```

# Quick data checks

```{r message=FALSE, warning=FALSE, results='hide'}
head(mpg)
glimpse(mpg)
summary(mpg)
```

# 2.17 Create your first ggplot

Using data from mpg, found in the ggplot package.

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

This plot shows the relationship between the engine size and the fuel efficiency.
Visually representing that cars from 1999 - 2008 with larger engines typically have lower fuel efficiency.

The plot shows a negative relationship between engine size (displ) and fuel efficiency (hwy).
In other words, cars with big engines use more fuel.
What does this say about fuel efficiency and engine size?

# 2.18 Understanding the 'grammar of graphics"

What happens if we just try to run ggplot on its own?

```{r message=FALSE, warning=FALSE, eval=FALSE}
ggplot() 
# Or with just the data specified?
ggplot(data = mpg)
```

We need to map some aesthetics!!
It is blank.

## 2.18.1 Graphing template

In all cases, ggplot code will follow a generic template like below:

```{r message=FALSE, warning=FALSE, eval=FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

When you’re creating a plot, you essentially need two attributes of a plot: a geom and aesthetics.

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

## 2.18.2 Aesthetic mappings

### Change point colour by class:

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```

This allows to see patterns within types of cars, as each class of car is a different color.

### Change point size by class:

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

Note the warning!!!
Instead of using color to differentiate between classes, this instead uses 'size', which makes it difficult to distinguish classes from one another.

### Change transparency (alpha) by class

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```

Another warning!!
**Question: When would using alpha (or size be appropriate?)** Alpha would be better used in a dense point cloud where you can see clusters a bit better.
Size is better used on a non-discrete variable when looking at a more continuous scale.

### Change point shape by class:

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

Warning: The shape palette can deal with a maximum of 6 discrete values because more than 6 becomes difficult to discriminate.
Consider specifying shapes manually if you need that many of them.

### Make all points blue

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

#### Aestheic something other than variable name

What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ \< 5)?
Note, you’ll also need to specify x and y.

``` {rmessage="FALSE," warning="FALSE"}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = displ < 5))
```

This creates everything after 5 as 'FALSE'.

# 2.19 Troubleshooting

Despite this, there will be some other common problems you’re likely to face.
Perhaps the main one with ggplot is having the + in the wrong place.
It has to go at the end of the line (not the start of the next line), which is particularly important if you’re running your script line-by-line.

Example:

```{r message=FALSE, warning=FALSE, eval=FALSE}
ggplot(data = mpg) 
+ geom_point(mapping = aes(x = displ, y = hwy)) 
# the + should be on top line
```

As with all programming, there are some really useful methods for debugging your code.
Revisit the Module 1 workshop where we discussed using online help, reading the R help pages, learning to read your own code carefully and precisely, and trying to interpret the error messages you’ve received.
As a very last resort, you can always ask the instructor (who will probably just tell you to do the above anyway).

# 2.20 Faceting

Faceting allows you to split your data into subsets.
AKA can have multiple panels in a figure.

To facet your plot using a single variable, use facet_wrap().
Facet wrap syntax is in the function of a formula (kind of like a linear model formula), where the \~ dictates which variable you want to subset your data with.
Note: only use facet_wrap() for discrete variables.

## facet_wrap

Split the plot out by car type (class)

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

## facet_grid

If you want to do this with more than one variable, then use facet_grid().
Here you need two variables using \~ to split them up.
Facet plot on the combination of two variables using facet_grid().
Facet grid needs two variables separated by a \~.

A separate facet for each combination of drive-type (e.g. 4WD) \* number of cylinders

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

Note that there are no occurrences of 5 cylinder 4WDs OR RWD vehicles!

## facet_grid by just row (or column)

Use a .
if you do not want to facet in the rows or column dimension.
Note what happens in this case:

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(cyl ~ .)
  #facet_grid(. ~ cyl) # Alternatively
```

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  #facet_grid(cyl ~ .)
  facet_grid(. ~ cyl) # Alternatively
```

# Exercise 1:

Read ?f
acet_wrap.
What does nrow do?
What does ncol do?
What other options control the layout of the individual panels?

```{r message=FALSE, warning=FALSE}
# ?facet_wrap
```

nrow, ncol -\> Number of rows and columns, respectively.
The other options that control the layout of individual panels are: - scales: control how axes behave across panels, and can be useful when panel values differ a lot.
- space: controls whether panel change size depending on their scale - dir: controls how panels fill across the page, and where the filling starts - strip.position: controls where the strip labels appear - switch: moves facet labels to opposite axis - drop: alots to keep or removed unused panels - axis: controls whether axis lines are drawn in interior panels - axis.labels: controls whether the numbers/text labels appear on the interior axes

# 2.21 Fitting simple lines

We already used points:

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

However, ggplot2 can use a variety of geom objects to represent the data.
Here, we might want to use bar plots, line charts, boxplots and so on.
Well we can handle this issue in ggplot directly using a different geom to plot the same data.
Here, instead of plotting points, we will use a smooth line.

## To display the same data as a smooth line

fit through the points use geom_smooth()

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_line(mapping = aes(x = displ, y = hwy))
```

**Question: why don't we use geom_line() here? What would that look like?** geom_line connects the observed data points in the order of the variable on the x-axis.
It The overall trend may be harder to see with the noise.
geom_smooth on the other hand fits a model to the data, displaying the smooth trend rather than the raw data points.

This graph displays both on top of each other.

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_line() +
  geom_smooth()
```

A geom is an object that your plot uses to represent the data.
To change the geom type in your plot, simply change the geom function that you add to your plot template.
Sometimes you may want to try a few things out, in which case you could use comments to help you remember what worked and what didn’t.

## Using comments (\#)

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  #geom_point(mapping = aes(x = displ, y = hwy)) + # points horrible 
  geom_smooth(mapping = aes(x = displ, y = hwy)) # try smooth line
```

**Question: how does R work out where to plot the line?** Can use the chunk output to inform us.
Can also use the help menu.
For the graph above, using geom_smooth, R used the model 'loess' using the formula y \~ x.
This method is often used for datasets with less than 1,000 observations.
The method (e.g. 'loess') is chosen based on the size of the largest group across all data panels.
'loess' also uses a t-based approximation for the standard error calculation.

## Changing linetype

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv)) # Can also use "lty = " for the same result
```

## Grouping

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```

## Change line colour based on drv value

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE,
  )
```

## Multiple geoms

Here we will plot multiple geoms on the single plot.
All you need to do is to add them together.
This one is nice for showing the underlying data and how it relates to the geom_smooth line.

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

A better way...

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```

By placing the mapping in the 'ggplot' portion of the code, ggplot allows you to pass these ‘overarching’ mappings to the ggplot() argument, making them global mappings that are applied to every single subsequent geom.
Much in the same way that the data argument is also global, it is used every time a geom is called.

## Can still manipulate each geom/layer separately:

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

Now we have different colours to indicate the raw data (geom_point) and one colour for the overall relationship (geom_smooth).
Neat!

You also can specify different data for various layers depending on what you are trying to display.

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

This graph specified the smooth line for just the class of subcompact cars.

**Question: what does the "se = FALSE" do?** 'se' displays the standard error confidence band around the smooth line.
By switching it to 'FALSE' that removes the band.

# Exercise 2:

**1. What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?** - Line chart: geom_line - Boxplot: geom_boxplot - Histogram:geom_histogram - Area chart: geom_area

**2. Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.**

```{r message=FALSE, warning=FALSE}
# Line chart
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_line()

# Boxplot
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_boxplot(mapping = aes(x = displ, y = hwy, group = drv))

# Histogram
ggplot(data = mpg, mapping = aes(x = displ)) + 
  geom_histogram(binwidth = 0.1)

# Area chart
ggplot(mpg, aes(x = displ, y = hwy, fill = class)) +
  geom_area()
```

I was unable to use the same exact variables for every one of these types of 'geom' as some are better suited to categorical vs. continuous data or one vs. two variables.

**3. Will these two graphs look different? Why/why not?**

```{r message=FALSE, warning=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

No, these will not look different.
In the first graph the data and mapping are set into the overarching portion of the code that creates them as global mapping applied to all subsequent geoms.
Whereas the one below specifies the same mappings and data, just in each of the geoms used.

